---
title: "Herder Rangeland MARL — MADDPG, No PES and Stock Survival"
author: "Khorloo Batpurev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---


> **Learning-algorithm conversion note.** This file preserves the original ecological transition table and all non-learning model rules from its source Rmd. The learning layer has been replaced by discrete-action MADDPG with one decentralised actor and one centralised critic per herder, masked straight-through Gumbel-Softmax action selection, joint replay memory, and Polyak target networks.

# MARL for PES design

There is no PES payment in this scenario; but agents still earn income from livestock. 

This is a simplified version with 6 agents and 25cells(5x5). Each agent has 2 designated pastures that they switch between each iteration. In an emergency they are allowed to choose a third pasture,that is not a designated pasture of someone else's. Co-location has a 0.7 penalty, so there is incentive to avoid overlapping with other agents in the same cell(pasture). Because simulation would be expensive on a continuous herd size, the model for now has four stocking categories as actions.Agents can't change herd size dramatically between two iterations,and hence there is an adjacent-category constraint:an agent may stay in the same category or move by one category per iteration, but cannot jump directly from`<200`to`>600`or vice versa. There is a 5 year ecosystem memory with an exponential decay weight(last year's condition is the most influential on this year's). Information sharing is only for the purpose of estimating neighbourhood density, which then enables agents to avoid a co-location penalty. 

MADDPG assumptions: each herder has a decentralised actor that observes only its own local observation. Training uses one centralised critic per herder, each critic observing the joint observations and joint actions of all six agents. The four original stocking categories and the adjacent-category constraint are retained through a masked straight-through Gumbel-Softmax actor policy.

## Model definition

The ecological environment, movement rules, stocking categories, action-based PES payments, stock-survival function, co-location penalty and emergency costs are unchanged from the source Q-learning version. Only the learning architecture is replaced.

For each agent \(i\), the local observation is the same information used previously:

\[
o_{i,t} =
(
\text{current cell},
\text{rangeland condition},
\text{weather},
\text{neighbour density},
\text{previous action}
).
\]

The categorical components are one-hot encoded, so no information is added or removed relative to the previous state key.

Each agent has its own actor network

\[
\pi_i(o_i;\theta_i)
\]

which outputs four action logits. The same/adjacent stocking-category constraint is applied as a hard action mask. During training, a straight-through Gumbel-Softmax draw converts the masked logits to an approximately differentiable one-hot action:

\[
y_{i,k}
=
\frac{
\exp((z_{i,k}+g_{i,k})/\tau_G)
}{
\sum_j \exp((z_{i,j}+g_{i,j})/\tau_G)
},
\]

with infeasible actions masked before the softmax. The realised landscape simulation always receives one of the four original discrete stocking categories.

Training is centralised and execution is decentralised. Agent \(i\)'s critic receives all agents' observations and actions:

\[
Q_i(o_1,\ldots,o_N,a_1,\ldots,a_N;\phi_i),
\]

while the actor continues to use only \(o_i\). Critic targets are

\[
y_i =
r_i +
\gamma(1-d)
Q_i^{\mathrm{target}}
(o'_1,\ldots,o'_N,a'_1,\ldots,a'_N).
\]

Actor \(i\) is updated to maximise its centralised critic,

\[
J_i(\theta_i)
=
E[
Q_i(o_1,\ldots,o_N,a_1,\ldots,\pi_i(o_i),\ldots,a_N)
].
\]

Experience is stored in a joint replay buffer and sampled in mini-batches. Target actor and critic networks are updated by Polyak averaging. Direct livestock income is still adjusted by expected stock survival and the action-based PES reward equation is unchanged.

## Spatial rules

- There are six agents and 25 cells.
- Each agent has two unique designated cells.
- Under ordinary conditions, the agent alternates deterministically between its two designated cells.
- Emergency movement is activated only when:
    1. current weather is drought; and
    2. both designated cells are degraded.
- An emergency destination:
    - cannot be any agent's designated cell;
    - must be ecologically better than both designated cells;
    - incurs travel and social costs.
- Travel and social costs receive equal weight: 0.5 + 0.5.
- Social cost measures passage through other agents' designated cells on the route to the emergency destination.
- If multiple agents occupy the same destination cell, their rewards receive a co-location penalty.

## Information sharing


All agents observe a shared public ledger containing the current locations and current herd sizes of all other agents. This information is used to calculate neighbour density. The shared ledger is deliberately simple in this version.

## Ecological memory


Every cell carries its own persistent 5-step exponential moving average (EMA) of ecological degradation severity. The stochastic ecological signal is drawn from the CSV-derived transition kernel in Step 5, which now permits constrained recovery as well as deterioration. This includes unassigned cells: if an emergency herder grazed a cell several iterations earlier, that legacy remains in the cell EMA and fades exponentially rather than being reset when the cell is ungrazed. The newest transition outcome receives the largest weight. The EMA defines the categorical `rangeland_condition` observed by agents. If one simulation step represents one year, this is approximately a five-year ecological memory; if one step represents a season, it is a five-season memory.


# . Setup

```{r setup}
library(tidyverse)
library(purrr)
library(ggplot2)
library(torch)
set.seed(128)
torch_manual_seed(128)
#Grid
grid_nx<-5L
grid_ny<-5L
#Rangeland degradation severity from best to worst
r_states<-c("intact","good","poor","degraded")
degradation_score<-c(intact=0,good=1,poor=2,degraded=3)
#Four discrete stocking-rate actions
actions<-c("<200","200-400","400-600",">600")
#"none" is when there is no grazing in the cell
ecological_footprints<-c("none",actions)
#Representative herd size used for information sharing and combined pressure
action_herd_size<-c("<200"=100,"200-400"=300,"400-600"=500,">600"=700)
#Starting stocking categories as placeholders
initial_actions<-c(A1="<200",A2="200-400",A3="200-400",A4="400-600",A5="400-600",A6=">600")
#Direct herd income before survival adjustment
herd_income<-c("<200"=0.8,"200-400"=2.0,"400-600"=2.5,">600"=3.5)
#Action-based PES
pes_payment<-0
#Expected stock survival by current rangeland condition and stocking action
stock_survival<-matrix(
  c(
    0.98,0.95,0.94,0.93,#intact
    0.96,0.94,0.93,0.92,#good
    0.84,0.83,0.8,0.7,#poor
    0.83,0.8,0.7,0.6#degraded
  ),
  nrow=4,
  byrow=TRUE,
  dimnames=list(r_states,actions)
)
#Weather. This gets changed in the sensitivity test to different frequencies
weather_prob<-c(drought=0.30,normal=0.50,wet=0.20)

#MADDPG hyperparameters
gamma<-0.95
actor_lr<-1e-3
critic_lr<-2e-3
gumbel_tau<-1.00
target_tau<-0.01
batch_size<-64L
replay_capacity<-50000L
learning_starts<-64L
updates_per_step<-1L
actor_hidden_dim<-128L
critic_hidden_dim<-256L
#Co-location penalty
colocation_multiplier<-0.70
#Emergency movement costs
travel_cost_weight<-0.50
social_cost_weight<-0.50
emergency_cost_scale<-1.00
#Ecosystem memory
memory_window<-5L
ema_alpha<-2/(memory_window+1)
```

```{r test-setup, include=FALSE}
stopifnot(length(actions)==4)
stopifnot(identical(ecological_footprints,c("none",actions)))
stopifnot(all(initial_actions%in%actions))
stopifnot(abs(sum(weather_prob)-1)<1e-12)
stopifnot(gamma>=0&&gamma<1)
stopifnot(actor_lr>0)
stopifnot(critic_lr>0)
stopifnot(gumbel_tau>0)
stopifnot(target_tau>0&&target_tau<=1)
stopifnot(batch_size>=1)
stopifnot(replay_capacity>=batch_size)
stopifnot(learning_starts>=batch_size)
stopifnot(ema_alpha>0&&ema_alpha<1)
stopifnot(abs(travel_cost_weight+social_cost_weight-1)<1e-12)
stopifnot(identical(dim(stock_survival),c(4L,4L)))
stopifnot(all(stock_survival>=0&stock_survival<=1))

tibble(
  action=actions,
  herd_income=herd_income[actions],
  survival_intact=stock_survival["intact",actions],
  expected_livestock_income=herd_income[actions]*stock_survival["intact",actions],
  pes_payment=pes_payment[actions],
  total_on_intact=herd_income[actions]*stock_survival["intact",actions]+pes_payment[actions]
)
```

# . Create the 5 × 5 landscape

Cells are indexed from 1 to 25. All cells begin intact.

```{r create-grid, include=FALSE}
make_land_grid<-function() {
  expand_grid(
    y = 0:(grid_ny - 1),
    x = 0:(grid_nx - 1)
  ) %>%
    arrange(y, x) %>%
    mutate(
      cell_id = row_number(),
      rangeland_condition = "intact",
      condition_ema_score = unname(degradation_score["intact"])
    ) %>%
    select(cell_id, x, y, rangeland_condition, condition_ema_score)
}
land_grid_initial<-make_land_grid()
```

```{r test-grid, include=FALSE}
stopifnot(nrow(land_grid_initial) == 25)
stopifnot(n_distinct(land_grid_initial$cell_id) == 25)
ggplot(land_grid_initial, aes(x, y, label = cell_id)) +
  geom_tile(fill = "white", colour = "grey50") +
  geom_text() +
  coord_equal() +
  scale_y_reverse() +
  labs(title = "TEST: 5 × 5 cell indexing") +
  theme_minimal()
```

# . Assign two designated cells to each of six agents

The 12 designated cells are mutually exclusive for each agent. 

```{r designated-cells}
designated_cells<-tribble(
  ~agent_id, ~slot, ~cell_id,
  "A1", 1L,  1L,
  "A1", 2L,  7L,
  "A2", 1L,  5L,
  "A2", 2L,  9L,
  "A3", 1L, 21L,
  "A3", 2L, 17L,
  "A4", 1L, 25L,
  "A4", 2L, 19L,
  "A5", 1L, 11L,
  "A5", 2L, 12L,
  "A6", 1L, 14L,
  "A6", 2L, 15L
)
designated_lookup<-designated_cells %>%
  arrange(agent_id, slot) %>%
  group_by(agent_id) %>%
  summarise(
    cell_1 = first(cell_id),
    cell_2 = last(cell_id),
    .groups = "drop"
  )
designated_owner<-designated_cells %>%
  distinct(cell_id, agent_id)
unassigned_cells<-setdiff(
  land_grid_initial$cell_id,
  designated_cells$cell_id
)
```

```{r test-designated-cells, include=FALSE}
stopifnot(n_distinct(designated_cells$cell_id) == 12)
stopifnot(length(unassigned_cells) == 13)
land_grid_initial %>%
  left_join(designated_owner, by = "cell_id") %>%
  mutate(agent_id = replace_na(agent_id, "Unassigned")) %>%
  ggplot(aes(x, y, fill = agent_id, label = cell_id)) +
  geom_tile(colour = "white") +
  geom_text() +
  coord_equal() +
  scale_y_reverse() +
  labs(
    title = "TEST: designated and unassigned cells",
    fill = "Owner"
  ) +
  theme_minimal()
```

# . Weather

Weather is exogenous and shared by the whole landscape at each time step.

```{r weather}
draw_weather<-function(n = 1L) {
  sample(
    names(weather_prob),
    size = n,
    replace = TRUE,
    prob = weather_prob
  )
}
```

```{r test-weather, include=FALSE}
set.seed(123)
weather_test<-tibble(weather = draw_weather(10000))
weather_test %>%
  count(weather) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(weather, prop)) +
  geom_col() +
  geom_point(
    data = tibble(
      weather = names(weather_prob),
      prop = as.numeric(weather_prob)
    ),
    aes(weather, prop),
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(
    title = "TEST: empirical weather frequency",
    subtitle = "Bars are simulated; points are target probabilities",
    y = "Proportion"
  ) +
  theme_minimal()
```

# . Expert derived ecological transition model


The expert derived transition table contains joint relationship between climate, stocking category, and three pasture outcomes; `intact`,`good`,`degraded`. There is no directly elicited `poor` state. We derive `poor` state by applying the exponential moving average (EMA) function, which produces a continuous number between 0 and 3. `poor` is assigned for values between 1.5 and 2.5. 

Due to time limitations on experts' time and cognitive load, not all starting conditions for the full transition table was possible to be elicited in the field. Only the starting condition `intact` is fully elicited. The remaining starting conditions (good, poor and degraded) are therefore are derived from the `intact` starting condition values i.e. purely assumption based at this stage.

The missing current-condition rows are constructed as follows for each action x weather combination:

\[
\mathbf p_G=\frac{2}{3}\mathbf p_I+\frac{1}{3}\mathbf p_D,
\]

and

\[
\mathbf p_P^{raw}=\frac{1}{3}\mathbf p_G+\frac{2}{3}\mathbf p_D.
\]

For current `poor` and `degraded` conditions, a direct transition signal to `intact` is prohibited. Any raw `intact` mass in the `poor` interpolation is transferred to `good`. Thus recovery from `poor` or `degraded` occurs through `good` rather than jumping directly to `intact`.

For a currently degraded cell:

- under `wet` + `none`, degraded -> good recovery probability is 0.80;
- under `wet` + `<200`, recovery is 0.50;
- under `wet` + `>600`, recovery is 0.01;
- `wet` + `200-400` and `400-600` are interpolated using the empirical action gradient in `1-P(degraded)` from the CSV;
- under `normal` + `none`, recovery is 0.66;
- under `drought` + `none`, recovery is 0.01;
- under `normal` or `drought` with any positive grazing category, a degraded cell remains degraded with probability 1.

For `none`, which is also absent from the elicitation, we have also made assumptions for now as a placeholder; intact-start persistence anchors are 0.99 under `wet` and `normal` and 0.01 under `drought`. The remaining probability is split between `good` and `degraded` using the same-weather `<200` empirical good:degraded ratio.

```{r base-transition}
#Empirical joint weather x stocking transition probabilities. These are the mean of 140 individual estimates. 
climate_transition_source<-tribble(
  ~weather,~action,~intact,~good,~degraded,
  "wet","<200",0.8665,0.0963,0.0372,
  "wet","200-400",0.7425,0.1576,0.0999,
  "wet","400-600",0.5945,0.2278,0.1777,
  "wet",">600",0.4349,0.2689,0.2962,
  "normal","<200",0.7438,0.1837,0.0725,
  "normal","200-400",0.5856,0.2404,0.1740,
  "normal","400-600",0.3926,0.2906,0.3168,
  "normal",">600",0.2581,0.2811,0.4608,
  "drought","<200",0.4906,0.3077,0.2017,
  "drought","200-400",0.3481,0.3067,0.3452,
  "drought","400-600",0.2327,0.2730,0.4943,
  "drought",">600",0.1261,0.2255,0.6484
)
transition_signal_states<-c("intact","good","degraded")
make_prob_vector<-function(intact,good,degraded){
  p<-c(intact=intact,good=good,poor=0,degraded=degraded)
  if(any(!is.finite(p))||any(p<0)||any(p>1)||abs(sum(p)-1)>1e-10){
    stop("Invalid ecological transition probability vector.")
  }
  p
}
empirical_intact_prob<-function(action,weather){
  stopifnot(action%in%actions,weather%in%names(weather_prob))
  row<-climate_transition_source[
    climate_transition_source$action==action&
      climate_transition_source$weather==weather,
    ,drop=FALSE
  ]
  if(nrow(row)!=1L)stop("Expected one empirical action x weather row.")
  make_prob_vector(row$intact[[1]],row$good[[1]],row$degraded[[1]])
}
#Explicit no-grazing anchors for an intact current condition.
none_intact_persistence<-c(wet=0.99,normal=0.99,drought=0.01)
make_none_intact_prob<-function(weather){
  stopifnot(weather%in%names(weather_prob))
  p_intact<-unname(none_intact_persistence[[weather]])
  reference<-empirical_intact_prob("<200",weather)
  residual<-1-p_intact
  gd_total<-reference[["good"]]+reference[["degraded"]]
  make_prob_vector(
    intact=p_intact,
    good=residual*reference[["good"]]/gd_total,
    degraded=residual*reference[["degraded"]]/gd_total
  )
}
intact_start_prob<-function(action,weather){
  if(action=="none")make_none_intact_prob(weather) else empirical_intact_prob(action,weather)
}
#Wet recovery curve for a currently degraded grazed cell.
wet_recovery_shape<-climate_transition_source%>%
  filter(weather=="wet")%>%
  mutate(recovery_favourability=1-degraded)%>%
  arrange(match(action,actions))
wet_fav_low<-wet_recovery_shape$recovery_favourability[
  wet_recovery_shape$action==">600"
]
wet_fav_high<-wet_recovery_shape$recovery_favourability[
  wet_recovery_shape$action=="<200"
]
wet_recovery_shape<-wet_recovery_shape%>%
  mutate(
    shape=(recovery_favourability-wet_fav_low)/(wet_fav_high-wet_fav_low),
    degraded_to_good=0.01+shape*(0.50-0.01)
  )
degraded_none_recovery<-c(wet=0.80,normal=0.66,drought=0.01)
degraded_recovery_prob<-function(action,weather){
  stopifnot(action%in%ecological_footprints,weather%in%names(weather_prob))
  if(action=="none")return(unname(degraded_none_recovery[[weather]]))
  if(weather!="wet")return(0)
  row<-wet_recovery_shape[wet_recovery_shape$action==action,,drop=FALSE]
  if(nrow(row)!=1L)stop("Expected one wet recovery row.")
  row$degraded_to_good[[1]]
}
degraded_start_prob<-function(action,weather){
  recovery<-degraded_recovery_prob(action,weather)
  make_prob_vector(intact=0,good=recovery,degraded=1-recovery)
}
apply_recovery_constraint<-function(prob,current_condition){
  stopifnot(current_condition%in%r_states)
  if(current_condition%in%c("poor","degraded")){
    prob[["good"]]<-prob[["good"]]+prob[["intact"]]
    prob[["intact"]]<-0
  }
  if(any(prob<0)||any(prob>1)||abs(sum(prob)-1)>1e-10){
    stop("Recovery constraint produced an invalid probability vector.")
  }
  prob
}
make_transition_matrix<-function(action,weather){
  stopifnot(action%in%ecological_footprints,weather%in%names(weather_prob))
  p_intact<-intact_start_prob(action,weather)
  p_degraded<-apply_recovery_constraint(
    degraded_start_prob(action,weather),
    "degraded"
  )
  p_good<-(2/3)*p_intact+(1/3)*p_degraded
  p_poor_raw<-(1/3)*p_good+(2/3)*p_degraded
  p_poor<-apply_recovery_constraint(p_poor_raw,"poor")
  M<-rbind(
    intact=p_intact,
    good=p_good,
    poor=p_poor,
    degraded=p_degraded
  )
  M<-M[r_states,r_states,drop=FALSE]
  if(any(!is.finite(M))||any(M<0)||any(M>1)||any(abs(rowSums(M)-1)>1e-10)){
    stop("Invalid ecological transition matrix.")
  }
  M
}
base_transition<-setNames(
  lapply(ecological_footprints,function(a){
    setNames(
      lapply(names(weather_prob),function(w){make_transition_matrix(a,w)}),
      names(weather_prob)
    )
  }),
  ecological_footprints
)
sample_next_condition<-function(current_condition,action,weather){
  transition_matrix<-base_transition[[action]][[weather]]
  probabilities<-transition_matrix[current_condition,,drop=TRUE]
  sample(names(probabilities),size=1,prob=probabilities)
}
```

# . Five-step ecosystem memory

This is **cell-level ecological memory**. Every one of the 25 cells stores its own persistent EMA, including cells that are currently ungrazed or unassigned.

For the ecological update to time \(t+1\), first draw a stochastic transition outcome \(z_{t+1}\) from the base transition using \(C_t\), \(A_t\), and \(W_{t+1}\). Then update the cell EMA:

\[
E_{t+1} =
\alpha_E z_{t+1} +
(1-\alpha_E)E_t.
\]

The newest outcome receives weight \(\alpha_E\); older ecological history decays geometrically. The categorical `rangeland_condition` observed by the agents is then defined from the EMA thresholds. This is the mechanism that defines the rangeland-condition component already present in \(s\).

```{r ecosystem-memory}
condition_from_ema<-function(x){
  case_when(
    x<0.5~"intact",
    x<1.5~"good",
    x<2.5~"poor",
    TRUE~"degraded"
  )
}
update_cell_memory<-function(old_ema,transition_outcome,alpha=ema_alpha){
  outcome_score<-unname(degradation_score[[transition_outcome]])
  new_ema<-alpha*outcome_score+(1-alpha)*old_ema
  list(ema_score=new_ema,condition=condition_from_ema(new_ema))
}
```

```{r test-ecosystem-memory, include=FALSE}
outcomes<-c("degraded","good","good","good","good")
ema_value<-0
memory_test<-map_dfr(seq_along(outcomes),function(t){
  ans<-update_cell_memory(old_ema=ema_value,transition_outcome=outcomes[t])
  ema_value<<-ans$ema_score
  tibble(step=t,transition_outcome=outcomes[t],ema_score=ans$ema_score,condition=ans$condition)
})
stopifnot(memory_test$ema_score[1]>0)
stopifnot(all(memory_test$ema_score>0))
stopifnot(abs(ema_alpha-1/3)<1e-12)
memory_test
```

# . Agent objects, neural networks and replay memory

Each agent retains its ecological and livestock state as before. Each agent now has its own actor and centralised critic; target networks and a joint replay buffer are used for MADDPG training.

```{r create-agents}
create_agent<-function(
    agent_id,
    initial_action=initial_actions[[agent_id]]
){
  stopifnot(initial_action%in%actions)

  first_cell<-designated_lookup%>%
    filter(.data$agent_id==.env$agent_id)%>%
    pull(cell_1)

  list(
    id=agent_id,
    current_cell=first_cell,
    previous_action=initial_action,
    herd_size=unname(action_herd_size[[initial_action]])
  )
}

agents_initial<-setNames(
  lapply(
    names(initial_actions),
    function(id){
      create_agent(
        agent_id=id,
        initial_action=initial_actions[[id]]
      )
    }
  ),
  names(initial_actions)
)

actor_network<-nn_module(
  "ActorNetwork",
  initialize=function(obs_dim,action_dim,hidden_dim=128L){
    self$net<-nn_sequential(
      nn_linear(obs_dim,hidden_dim),
      nn_relu(),
      nn_linear(hidden_dim,hidden_dim),
      nn_relu(),
      nn_linear(hidden_dim,action_dim)
    )
  },
  forward=function(x){
    self$net(x)
  }
)

critic_network<-nn_module(
  "CentralisedCritic",
  initialize=function(joint_obs_dim,joint_action_dim,hidden_dim=256L){
    self$net<-nn_sequential(
      nn_linear(joint_obs_dim+joint_action_dim,hidden_dim),
      nn_relu(),
      nn_linear(hidden_dim,hidden_dim),
      nn_relu(),
      nn_linear(hidden_dim,1L)
    )
  },
  forward=function(joint_obs,joint_action){
    self$net(torch_cat(list(joint_obs,joint_action),dim=2))
  }
)

copy_module_state<-function(target,source){
  target$load_state_dict(source$state_dict())
  invisible(target)
}

new_replay_buffer<-function(capacity=replay_capacity){
  buffer<-new.env(parent=emptyenv())
  buffer$capacity<-as.integer(capacity)
  buffer$size<-0L
  buffer$position<-1L
  buffer$obs<-vector("list",buffer$capacity)
  buffer$actions<-vector("list",buffer$capacity)
  buffer$rewards<-vector("list",buffer$capacity)
  buffer$next_obs<-vector("list",buffer$capacity)
  buffer$masks<-vector("list",buffer$capacity)
  buffer$next_masks<-vector("list",buffer$capacity)
  buffer$done<-vector("list",buffer$capacity)
  buffer
}

replay_add<-function(
    buffer,
    obs,
    action_onehot,
    rewards,
    next_obs,
    masks,
    next_masks,
    done=0
){
  i<-buffer$position

  buffer$obs[[i]]<-as.vector(t(obs))
  buffer$actions[[i]]<-as.vector(t(action_onehot))
  buffer$rewards[[i]]<-as.numeric(rewards)
  buffer$next_obs[[i]]<-as.vector(t(next_obs))
  buffer$masks[[i]]<-as.vector(t(masks))
  buffer$next_masks[[i]]<-as.vector(t(next_masks))
  buffer$done[[i]]<-as.numeric(done)

  buffer$size<-min(buffer$size+1L,buffer$capacity)
  buffer$position<-if(i>=buffer$capacity)1L else i+1L

  invisible(buffer)
}

replay_sample<-function(buffer,n=batch_size){
  stopifnot(buffer$size>=n)
  idx<-sample.int(buffer$size,n,replace=FALSE)

  stack_field<-function(field){
    do.call(
      rbind,
      lapply(idx,function(i)buffer[[field]][[i]])
    )
  }

  list(
    obs=stack_field("obs"),
    actions=stack_field("actions"),
    rewards=stack_field("rewards"),
    next_obs=stack_field("next_obs"),
    masks=stack_field("masks"),
    next_masks=stack_field("next_masks"),
    done=matrix(
      unlist(lapply(idx,function(i)buffer$done[[i]])),
      ncol=1
    )
  )
}
```

```{r test-agents, include=FALSE}
stopifnot(length(agents_initial)==6)
stopifnot(all(vapply(agents_initial,is.list,logical(1))))
stopifnot(all(
  vapply(
    agents_initial,
    function(a)a$previous_action%in%actions,
    logical(1)
  )
))
stopifnot(all(
  vapply(
    agents_initial,
    function(a)a$herd_size>0,
    logical(1)
  )
))

replay_test<-new_replay_buffer(10L)
stopifnot(replay_test$size==0L)
```

# . Shared information for neighbourhood density estimation

The public information ledger contains:

- agent identity;
- current cell;
- coordinates;
- previous action;
- current herd size.

Neighbour density is the number of **other** agents in the neighbourhood of the focal agent (one-cell radius in x and y).

```{r information-sharing}
share_agent_information<-function(
    agents,
    land_grid
) {
  map_dfr(
    agents,
    function(agent) {
      loc<-land_grid %>%
        filter(cell_id == agent$current_cell)
      tibble(
        agent_id = agent$id,
        current_cell = agent$current_cell,
        x = loc$x,
        y = loc$y,
        previous_action = agent$previous_action,
        herd_size = agent$herd_size
      )
    }
  )
}
compute_neighbour_density<-function(
    agent_id,
    shared_info,
    radius = 1L
) {
  focal<-shared_info %>%
    filter(.data$agent_id==.env$agent_id)
  others<-shared_info %>%
    filter(.data$agent_id!=.env$agent_id)
  sum(
    abs(others$x - focal$x) <= radius &
      abs(others$y - focal$y) <= radius
  )
}
```

```{r test-information-sharing, include=FALSE}
shared_test<-share_agent_information(
  agents_initial,
  land_grid_initial
)
density_test<-shared_test %>%
  mutate(
    neighbour_density = map_int(
      agent_id,
      compute_neighbour_density,
      shared_info = shared_test
    )
  )
stopifnot(all(
  density_test$neighbour_density >= 0 &
    density_test$neighbour_density <= 5
))
ggplot(
  density_test,
  aes(x, y, size = herd_size, label = agent_id)
) +
  geom_point() +
  geom_text(nudge_y = -0.25) +
  coord_equal() +
  scale_y_reverse() +
  labs(
    title = "TEST: public information-sharing ledger",
    subtitle = "Point size represents currently observed herd size"
  ) +
  theme_minimal()
```

# . Local observation encoding

The exact state requested for policy learning is

\[
(
\text{current cell},
\text{rangeland condition},
\text{weather},
\text{neighbour density},
\text{previous action}
).
\]

```{r state-encoding}
encode_state<-function(
    current_cell,
    rangeland_condition,
    weather,
    neighbour_density,
    previous_action
){
  paste(
    paste0("cell=",current_cell),
    paste0("land=",rangeland_condition),
    paste0("weather=",weather),
    paste0("neigh=",neighbour_density),
    paste0("prev=",previous_action),
    sep="|"
  )
}

one_hot_numeric<-function(value,levels){
  as.numeric(levels==value)
}

observation_dim<-
  grid_nx*grid_ny+
  length(r_states)+
  length(weather_prob)+
  6L+
  length(actions)

encode_observation<-function(
    current_cell,
    rangeland_condition,
    weather,
    neighbour_density,
    previous_action
){
  c(
    one_hot_numeric(current_cell,seq_len(grid_nx*grid_ny)),
    one_hot_numeric(rangeland_condition,r_states),
    one_hot_numeric(weather,names(weather_prob)),
    one_hot_numeric(neighbour_density,0:5),
    one_hot_numeric(previous_action,actions)
  )
}

observe_agent_state<-function(
    agent,
    land_grid,
    weather,
    shared_info
){
  current_condition<-land_grid%>%
    filter(cell_id==agent$current_cell)%>%
    pull(rangeland_condition)

  neighbour_density<-compute_neighbour_density(
    agent_id=agent$id,
    shared_info=shared_info
  )

  encode_state(
    current_cell=agent$current_cell,
    rangeland_condition=current_condition,
    weather=weather,
    neighbour_density=neighbour_density,
    previous_action=agent$previous_action
  )
}

observe_agent_vector<-function(
    agent,
    land_grid,
    weather,
    shared_info
){
  current_condition<-land_grid%>%
    filter(cell_id==agent$current_cell)%>%
    pull(rangeland_condition)

  neighbour_density<-compute_neighbour_density(
    agent_id=agent$id,
    shared_info=shared_info
  )

  encode_observation(
    current_cell=agent$current_cell,
    rangeland_condition=current_condition,
    weather=weather,
    neighbour_density=neighbour_density,
    previous_action=agent$previous_action
  )
}

create_maddpg_system<-function(agent_ids=names(initial_actions)){
  n_agents<-length(agent_ids)
  action_dim<-length(actions)

  joint_obs_dim<-n_agents*observation_dim
  joint_action_dim<-n_agents*action_dim

  actors<-setNames(
    lapply(
      agent_ids,
      function(id){
        actor_network(
          obs_dim=observation_dim,
          action_dim=action_dim,
          hidden_dim=actor_hidden_dim
        )
      }
    ),
    agent_ids
  )

  target_actors<-setNames(
    lapply(
      agent_ids,
      function(id){
        m<-actor_network(
          obs_dim=observation_dim,
          action_dim=action_dim,
          hidden_dim=actor_hidden_dim
        )
        copy_module_state(m,actors[[id]])
        m
      }
    ),
    agent_ids
  )

  critics<-setNames(
    lapply(
      agent_ids,
      function(id){
        critic_network(
          joint_obs_dim=joint_obs_dim,
          joint_action_dim=joint_action_dim,
          hidden_dim=critic_hidden_dim
        )
      }
    ),
    agent_ids
  )

  target_critics<-setNames(
    lapply(
      agent_ids,
      function(id){
        m<-critic_network(
          joint_obs_dim=joint_obs_dim,
          joint_action_dim=joint_action_dim,
          hidden_dim=critic_hidden_dim
        )
        copy_module_state(m,critics[[id]])
        m
      }
    ),
    agent_ids
  )

  actor_optimizers<-setNames(
    lapply(
      actors,
      function(m)optim_adam(m$parameters,lr=actor_lr)
    ),
    agent_ids
  )

  critic_optimizers<-setNames(
    lapply(
      critics,
      function(m)optim_adam(m$parameters,lr=critic_lr)
    ),
    agent_ids
  )

  list(
    agent_ids=agent_ids,
    actors=actors,
    target_actors=target_actors,
    critics=critics,
    target_critics=target_critics,
    actor_optimizers=actor_optimizers,
    critic_optimizers=critic_optimizers,
    updates=0L
  )
}
```

```{r test-state-encoding, include=FALSE}
state_test<-map_chr(
  agents_initial,
  observe_agent_state,
  land_grid=land_grid_initial,
  weather="normal",
  shared_info=shared_test
)

obs_test<-map(
  agents_initial,
  observe_agent_vector,
  land_grid=land_grid_initial,
  weather="normal",
  shared_info=shared_test
)

stopifnot(length(state_test)==6)
stopifnot(all(str_detect(state_test,"cell=")))
stopifnot(all(str_detect(state_test,"land=")))
stopifnot(all(str_detect(state_test,"weather=")))
stopifnot(all(str_detect(state_test,"neigh=")))
stopifnot(all(str_detect(state_test,"prev=")))
stopifnot(all(map_int(obs_test,length)==observation_dim))
stopifnot(all(map_dbl(obs_test,sum)==5))

maddpg_test<-create_maddpg_system()
stopifnot(length(maddpg_test$actors)==6)
stopifnot(length(maddpg_test$critics)==6)
```

# . Gumbel-Softmax actor with stocking-category anchoring

Action choice is based only on Q values for the currently observed state, but the feasible action set is constrained by the previous stocking category. The agent may stay in the same category or move one category up/down in one iteration. This is the discrete approximation to the stocking-change constraint. 

```{r gumbel-policy}
allowed_actions<-function(previous_action){
  switch(
    previous_action,
    "<200"=c("<200","200-400"),
    "200-400"=c("<200","200-400","400-600"),
    "400-600"=c("200-400","400-600",">600"),
    ">600"=c("400-600",">600"),
    stop("Unknown previous action: ",previous_action)
  )
}

action_mask<-function(previous_action){
  as.numeric(actions%in%allowed_actions(previous_action))
}

mask_logits_tensor<-function(logits,mask_tensor){
  logits+(1-mask_tensor)*(-1e9)
}

select_action_actor<-function(
    actor,
    observation,
    previous_action,
    explore=TRUE,
    temperature=gumbel_tau
){
  stopifnot(length(observation)==observation_dim)
  stopifnot(previous_action%in%actions)

  obs_tensor<-torch_tensor(
    matrix(observation,nrow=1),
    dtype=torch_float()
  )

  mask_vec<-action_mask(previous_action)

  mask_tensor<-torch_tensor(
    matrix(mask_vec,nrow=1),
    dtype=torch_float()
  )

  actor$eval()

  out<-with_no_grad({
    logits<-actor(obs_tensor)
    masked_logits<-mask_logits_tensor(logits,mask_tensor)
    probabilities<-nnf_softmax(masked_logits,dim=2)

    selected<-if(explore){
      nnf_gumbel_softmax(
        masked_logits,
        tau=temperature,
        hard=TRUE,
        dim=2
      )
    }else{
      probabilities
    }

    list(
      probabilities=probabilities,
      selected=selected
    )
  })

  probability_vector<-as.numeric(
    as_array(out$probabilities$cpu())
  )

  selected_vector<-as.numeric(
    as_array(out$selected$cpu())
  )

  chosen_index<-which.max(selected_vector)
  chosen_action<-actions[[chosen_index]]

  hard_one_hot<-rep(0,length(actions))
  hard_one_hot[[chosen_index]]<-1

  full_probabilities<-setNames(
    probability_vector,
    actions
  )

  list(
    action=chosen_action,
    action_index=chosen_index,
    one_hot=hard_one_hot,
    feasible_actions=allowed_actions(previous_action),
    probabilities=full_probabilities
  )
}
```

```{r test-gumbel-policy, include=FALSE}
stopifnot(identical(
  allowed_actions("<200"),
  c("<200","200-400")
))
stopifnot(!">600"%in%allowed_actions("<200"))
stopifnot(!"<200"%in%allowed_actions(">600"))

torch_manual_seed(99)
policy_system_test<-create_maddpg_system()
policy_obs_test<-observe_agent_vector(
  agents_initial$A1,
  land_grid_initial,
  "normal",
  shared_test
)

policy_result_test<-select_action_actor(
  actor=policy_system_test$actors$A1,
  observation=policy_obs_test,
  previous_action="<200",
  explore=TRUE
)

stopifnot(policy_result_test$action%in%c("<200","200-400"))
stopifnot(policy_result_test$probabilities[[">600"]]<1e-8)
stopifnot(abs(sum(policy_result_test$probabilities)-1)<1e-6)
```

# . Normal movement between two designated cells

Normal location is deterministic: agents alternate between their two designated cells.

```{r normal-movement}
normal_cell_for_step<-function(
    agent_id,
    step
) {
  row<-designated_lookup %>%
    filter(.data$agent_id==.env$agent_id)
  if (step %% 2L == 1L) {
    row$cell_1
  } else {
    row$cell_2
  }
}
```

```{r test-normal-movement, include=FALSE}
expand_grid(
  agent_id = paste0("A", 1:6),
  step = 1:6
) %>%
  mutate(
    cell_id = map2_int(
      agent_id,
      step,
      normal_cell_for_step
    )
  ) %>%
  ggplot(aes(step, cell_id, group = agent_id, colour = agent_id)) +
  geom_line() +
  geom_point() +
  labs(
    title = "TEST: deterministic alternation between two designated cells",
    y = "Cell ID"
  ) +
  theme_minimal()
```

# . Emergency trigger and emergency destination

The emergency gate is intentionally simple:

\[
\text{Emergency}_{i,t}
=
\mathbf{1}
(
W_t = \text{drought}
\land
C_{i,1} = \text{degraded}
\land
C_{i,2} = \text{degraded}
).
\]

There is no stochastic mobility propensity in this basic version.

## .1 Straight-line route through the grid

The route approximation is used only to calculate social trespass cost.

```{r line-path}
cells_on_straight_line<-function(
    origin_cell,
    destination_cell,
    land_grid,
    n_points = 101L
) {
  origin<-land_grid %>%
    filter(cell_id == origin_cell)
  destination<-land_grid %>%
    filter(cell_id == destination_cell)
  xs<-seq(origin$x, destination$x, length.out = n_points)
  ys<-seq(origin$y, destination$y, length.out = n_points)
  tibble(
    x = round(xs),
    y = round(ys)
  ) %>%
    distinct(x, y) %>%
    left_join(
      land_grid %>% select(cell_id, x, y),
      by = c("x", "y")
    ) %>%
    filter(!is.na(cell_id)) %>%
    pull(cell_id)
}
```

```{r test-line-path, include=FALSE}
line_test<-cells_on_straight_line(
  origin_cell = 1,
  destination_cell = 25,
  land_grid = land_grid_initial
)
land_grid_initial %>%
  mutate(on_path = cell_id %in% line_test) %>%
  ggplot(aes(x, y, fill = on_path, label = cell_id)) +
  geom_tile(colour = "white") +
  geom_text() +
  coord_equal() +
  scale_y_reverse() +
  labs(
    title = "TEST: cells intersected by a straight-line emergency route",
    fill = "On path"
  ) +
  theme_minimal()
```

## .2 Travel and social costs

Travel cost is normalized Euclidean distance.Social cost is the fraction of intermediate route cells that are designated to **another** agent. The destination itself must be unassigned, so social cost comes from crossing other people's pastures rather than occupying their designated destination.

```{r emergency-cost}
emergency_movement_cost<-function(
    agent_id,
    origin_cell,
    destination_cell,
    land_grid
) {
  origin<-land_grid %>%
    filter(cell_id == origin_cell)
  destination<-land_grid %>%
    filter(cell_id == destination_cell)
  distance<-sqrt(
    (origin$x - destination$x)^2 +
      (origin$y - destination$y)^2
  )
  max_distance<-sqrt(
    (grid_nx - 1)^2 +
      (grid_ny - 1)^2
  )
  travel_cost<-distance / max_distance
  path_cells<-cells_on_straight_line(
    origin_cell,
    destination_cell,
    land_grid
  )
  intermediate_cells<-setdiff(
    path_cells,
    c(origin_cell, destination_cell)
  )
  other_designated<-designated_cells %>%
    filter(.data$agent_id!=.env$agent_id) %>%
    pull(cell_id) %>%
    unique()
  social_cost<-if (
    length(intermediate_cells) == 0
  ) {
    0
  } else {
    mean(
      intermediate_cells %in% other_designated
    )
  }
  total_cost<-travel_cost_weight*travel_cost+social_cost_weight*social_cost
  list(
    travel_cost = travel_cost,
    social_cost = social_cost,
    total_cost = total_cost,
    path_cells = path_cells
  )
}
```

```{r test-emergency-cost, include=FALSE}
cost_test<-map_dfr(
  unassigned_cells,
  function(dest) {
    cost<-emergency_movement_cost(
      agent_id = "A1",
      origin_cell = 1,
      destination_cell = dest,
      land_grid = land_grid_initial
    )
    tibble(
      destination = dest,
      travel = cost$travel_cost,
      social = cost$social_cost,
      total = cost$total_cost
    )
  }
)
cost_test %>%
  pivot_longer(
    c(travel, social),
    names_to = "cost_type",
    values_to = "cost"
  ) %>%
  ggplot(aes(factor(destination), cost, fill = cost_type)) +
  geom_col() +
  labs(
    title = "TEST: 50/50 travel and social components",
    x = "Candidate emergency cell",
    y = "Unweighted component"
  ) +
  theme_minimal()
```

## .3 Emergency candidate set

An emergency destination must be:

- be unassigned cell that is not a designated cell for another agent;
- be better than both of the agent's own designated cells;
- therefore be non-degraded when both designated cells are degraded.

Among valid candidates, the basic version chooses the cell with the lowest combined emergency cost.

```{r emergency-trigger}
emergency_active<-function(
    agent_id,
    weather,
    land_grid
) {
  if (weather != "drought") {
    return(FALSE)
  }
  designated<-designated_cells %>%
    filter(.data$agent_id==.env$agent_id) %>%
    pull(cell_id)
  conditions<-land_grid %>%
    filter(cell_id %in% designated) %>%
    arrange(match(cell_id, designated)) %>%
    pull(rangeland_condition)
  length(conditions) == 2 &&
    all(conditions == "degraded")
}
select_emergency_cell<-function(
    agent_id,
    origin_cell,
    land_grid
) {
  designated<-designated_cells %>%
    filter(.data$agent_id==.env$agent_id) %>%
    pull(cell_id)
  designated_scores<-land_grid %>%
    filter(cell_id %in% designated) %>%
    pull(rangeland_condition) %>%
    degradation_score[.]
  #Candidate must be strictly better than BOTH designated cells.
  max_allowed_score<-min(designated_scores)
  candidates<-land_grid %>%
    filter(
      cell_id %in% unassigned_cells,
      degradation_score[rangeland_condition] < max_allowed_score
    )
  if (nrow(candidates) == 0) {
    return(NULL)
  }
  candidate_costs<-map_dfr(
    candidates$cell_id,
    function(destination_cell) {
      cost<-emergency_movement_cost(
        agent_id = agent_id,
        origin_cell = origin_cell,
        destination_cell = destination_cell,
        land_grid = land_grid
      )
      tibble(
        destination_cell = destination_cell,
        travel_cost = cost$travel_cost,
        social_cost = cost$social_cost,
        total_cost = cost$total_cost
      )
    }
  )
  best<-candidate_costs %>%
    filter(total_cost == min(total_cost)) %>%
    slice_sample(n = 1)
  list(
    cell_id = best$destination_cell,
    travel_cost = best$travel_cost,
    social_cost = best$social_cost,
    total_cost = best$total_cost,
    all_candidates = candidate_costs
  )
}
resolve_location<-function(
    agent_id,
    step,
    weather,
    land_grid
) {
  normal_cell<-normal_cell_for_step(
    agent_id,
    step
  )
  if (!emergency_active(
    agent_id,
    weather,
    land_grid
  )) {
    return(list(
      cell_id = normal_cell,
      movement_mode = "designated",
      emergency_cost = 0,
      travel_cost = 0,
      social_cost = 0
    ))
  }
  emergency<-select_emergency_cell(
    agent_id = agent_id,
    origin_cell = normal_cell,
    land_grid = land_grid
  )
  if (is.null(emergency)) {
    return(list(
      cell_id = normal_cell,
      movement_mode = "emergency_no_destination",
      emergency_cost = 0,
      travel_cost = 0,
      social_cost = 0
    ))
  }
  list(
    cell_id = emergency$cell_id,
    movement_mode = "emergency",
    emergency_cost = emergency$total_cost,
    travel_cost = emergency$travel_cost,
    social_cost = emergency$social_cost
  )
}
```

```{r test-emergency-trigger, include=FALSE}
emergency_grid_test<-land_grid_initial
a1_cells<-designated_cells %>%
  filter(agent_id == "A1") %>%
  pull(cell_id)
emergency_grid_test$rangeland_condition[
  emergency_grid_test$cell_id %in% a1_cells
]<-"degraded"
emergency_grid_test$condition_ema_score[
  emergency_grid_test$cell_id %in% a1_cells
]<-3
stopifnot(
  emergency_active(
    "A1",
    "drought",
    emergency_grid_test
  )
)
stopifnot(
  !emergency_active(
    "A1",
    "normal",
    emergency_grid_test
  )
)
emergency_demo<-resolve_location(
  agent_id = "A1",
  step = 1,
  weather = "drought",
  land_grid = emergency_grid_test
)
stopifnot(
  emergency_demo$cell_id %in% unassigned_cells
)
stopifnot(
  emergency_grid_test$rangeland_condition[
    emergency_grid_test$cell_id == emergency_demo$cell_id
  ] != "degraded"
)
land_grid_initial %>%
  mutate(
    role = case_when(
      cell_id %in% a1_cells ~ "A1 designated (degraded)",
      cell_id == emergency_demo$cell_id ~ "Selected emergency cell",
      TRUE ~ "Other"
    )
  ) %>%
  ggplot(aes(x, y, fill = role, label = cell_id)) +
  geom_tile(colour = "white") +
  geom_text() +
  coord_equal() +
  scale_y_reverse() +
  labs(
    title = "TEST: drought + two degraded designated cells activates emergency movement"
  ) +
  theme_minimal()
```

# . Action-based PES reward

Immediate reward contains:

1. herd income multiplied by expected stock survival for the current rangeland condition and stocking action;
2. action-based PES payment;
3. co-location penalty;
4. emergency movement cost when applicable.

Stock survival is the only multiplier on direct livestock income in this baseline. Stock survival affects current-period reward, while the intended herd size selected by the action still determines current grazing pressure. Surviving herd size is recorded diagnostically but is not yet carried forward as an endogenous herd-state variable.

Importantly, this reward enters the MADDPG update directly, so it influences future policy.

```{r reward}
get_stock_survival<-function(action,rangeland_condition){
  unname(stock_survival[rangeland_condition,action])
}
compute_reward<-function(action,rangeland_condition,n_agents_same_cell=1L,emergency_cost=0){
  survival_rate<-get_stock_survival(action=action,rangeland_condition=rangeland_condition)
  intended_herd_size<-unname(action_herd_size[[action]])
  expected_surviving_herd<-intended_herd_size*survival_rate
  expected_stock_loss<-intended_herd_size-expected_surviving_herd
  livestock_income<-herd_income[[action]]*survival_rate
  pes_reward<-pes_payment[[action]]
  gross_reward<-livestock_income+pes_reward
  interaction_multiplier<-colocation_multiplier^max(n_agents_same_cell-1L,0L)
  reward_after_interaction<-gross_reward*interaction_multiplier
  net_reward<-reward_after_interaction-emergency_cost_scale*emergency_cost
  list(
    net_reward=net_reward,
    livestock_income=livestock_income,
    pes_reward=pes_reward,
    gross_reward=gross_reward,
    interaction_multiplier=interaction_multiplier,
    emergency_cost=emergency_cost,
    stock_survival=survival_rate,
    intended_herd_size=intended_herd_size,
    expected_surviving_herd=expected_surviving_herd,
    expected_stock_loss=expected_stock_loss
  )
}
```

```{r test-reward, include=FALSE}
reward_test<-expand_grid(action=actions,condition=r_states)%>%
  mutate(
    stock_survival=map2_dbl(action,condition,~get_stock_survival(.x,.y)),
    reward=map2_dbl(action,condition,~compute_reward(action=.x,rangeland_condition=.y)$net_reward)
  )
stopifnot(all(reward_test$stock_survival>=0&reward_test$stock_survival<=1))
stopifnot(all(map_dbl(actions,~compute_reward(.x,"degraded")$livestock_income)<herd_income[actions]))
reward_test%>%
  ggplot(aes(action,reward,fill=condition))+
  geom_col(position="dodge")+
  labs(
    title="TEST: immediate reward enters learning",
    subtitle="Rangeland condition affects livestock income through stock survival",
    y="Net reward before movement/interaction cost"
  )+
  theme_minimal()
```

# . MADDPG actor-critic update

MADDPG updates one centralised critic and one decentralised actor per agent from mini-batches sampled from the joint replay buffer.

```{r maddpg-training}
agent_obs_cols<-function(i){
  ((i-1L)*observation_dim+1L):(i*observation_dim)
}

agent_action_cols<-function(i){
  action_dim<-length(actions)
  ((i-1L)*action_dim+1L):(i*action_dim)
}

soft_update_module<-function(target,source,tau_value=target_tau){
  target_parameters<-target$parameters
  source_parameters<-source$parameters

  stopifnot(length(target_parameters)==length(source_parameters))

  with_no_grad({
    for(k in seq_along(target_parameters)){
      target_parameters[[k]]$mul_(1-tau_value)
      target_parameters[[k]]$add_(
        source_parameters[[k]],
        alpha=tau_value
      )
    }
  })

  invisible(target)
}

build_joint_policy_actions<-function(
    system,
    joint_obs_tensor,
    joint_mask_tensor,
    focal_agent=NULL,
    target=FALSE
){
  action_tensors<-vector("list",length(system$agent_ids))

  for(j in seq_along(system$agent_ids)){
    id<-system$agent_ids[[j]]
    obs_j<-joint_obs_tensor[,agent_obs_cols(j)]
    mask_j<-joint_mask_tensor[,agent_action_cols(j)]

    actor<-if(target){
      system$target_actors[[id]]
    }else{
      system$actors[[id]]
    }

    logits<-actor(obs_j)
    masked_logits<-mask_logits_tensor(logits,mask_j)

    if(target){
      # Target actions are deterministic soft probability vectors.
      # The critic accepts continuous action vectors during learning.
      action_j<-nnf_softmax(masked_logits,dim=2)
    }else if(identical(id,focal_agent)){
      # Straight-through Gumbel-Softmax: hard in the forward pass,
      # differentiable in the backward pass.
      action_j<-nnf_gumbel_softmax(
        masked_logits,
        tau=gumbel_tau,
        hard=TRUE,
        dim=2
      )
    }else{
      # Other agents are treated as fixed while actor i is updated.
      action_j<-nnf_softmax(masked_logits,dim=2)$detach()
    }

    action_tensors[[j]]<-action_j
  }

  torch_cat(action_tensors,dim=2)
}

maddpg_train_step<-function(system,replay_buffer){
  if(replay_buffer$size<learning_starts){
    return(
      list(
        system=system,
        log=tibble()
      )
    )
  }

  batch<-replay_sample(
    replay_buffer,
    n=batch_size
  )

  obs_tensor<-torch_tensor(batch$obs,dtype=torch_float())
  actions_tensor<-torch_tensor(batch$actions,dtype=torch_float())
  rewards_tensor<-torch_tensor(batch$rewards,dtype=torch_float())
  next_obs_tensor<-torch_tensor(batch$next_obs,dtype=torch_float())
  masks_tensor<-torch_tensor(batch$masks,dtype=torch_float())
  next_masks_tensor<-torch_tensor(batch$next_masks,dtype=torch_float())
  done_tensor<-torch_tensor(batch$done,dtype=torch_float())

  loss_rows<-vector("list",length(system$agent_ids))

  # ---------------------------
  # 1. Centralised critic update
  # ---------------------------
  target_joint_actions<-with_no_grad({
    build_joint_policy_actions(
      system=system,
      joint_obs_tensor=next_obs_tensor,
      joint_mask_tensor=next_masks_tensor,
      target=TRUE
    )
  })

  critic_losses<-numeric(length(system$agent_ids))

  for(i in seq_along(system$agent_ids)){
    id<-system$agent_ids[[i]]

    system$critics[[id]]$train()
    system$critic_optimizers[[id]]$zero_grad()

    q_pred<-system$critics[[id]](
      obs_tensor,
      actions_tensor
    )

    reward_i<-rewards_tensor[,i]$unsqueeze(2)

    q_next<-with_no_grad({
      system$target_critics[[id]](
        next_obs_tensor,
        target_joint_actions
      )
    })

    q_target<-reward_i+
      gamma*(1-done_tensor)*q_next

    critic_loss<-nnf_mse_loss(
      q_pred,
      q_target
    )

    critic_loss$backward()
    system$critic_optimizers[[id]]$step()

    critic_losses[[i]]<-critic_loss$item()
  }

  # ---------------------------
  # 2. Decentralised actor update
  # ---------------------------
  actor_losses<-numeric(length(system$agent_ids))

  for(i in seq_along(system$agent_ids)){
    id<-system$agent_ids[[i]]

    system$actors[[id]]$train()
    system$actor_optimizers[[id]]$zero_grad()
    system$critic_optimizers[[id]]$zero_grad()

    policy_joint_actions<-build_joint_policy_actions(
      system=system,
      joint_obs_tensor=obs_tensor,
      joint_mask_tensor=masks_tensor,
      focal_agent=id,
      target=FALSE
    )

    actor_loss<-
      -system$critics[[id]](
        obs_tensor,
        policy_joint_actions
      )$mean()

    actor_loss$backward()
    system$actor_optimizers[[id]]$step()

    actor_losses[[i]]<-actor_loss$item()
  }

  # ---------------------------
  # 3. Polyak target-network update
  # ---------------------------
  for(id in system$agent_ids){
    soft_update_module(
      system$target_actors[[id]],
      system$actors[[id]]
    )
    soft_update_module(
      system$target_critics[[id]],
      system$critics[[id]]
    )
  }

  system$updates<-system$updates+1L

  log<-tibble(
    update=system$updates,
    agent_id=system$agent_ids,
    actor_loss=actor_losses,
    critic_loss=critic_losses
  )

  list(
    system=system,
    log=log
  )
}
```

```{r test-maddpg-training, include=FALSE}
training_system_test<-create_maddpg_system()
training_buffer_test<-new_replay_buffer(100L)

dummy_obs<-matrix(
  0,
  nrow=length(initial_actions),
  ncol=observation_dim
)
dummy_obs[,1]<-1
dummy_action<-diag(length(actions))[
  rep(1,length(initial_actions)),
  ,
  drop=FALSE
]
dummy_masks<-matrix(
  1,
  nrow=length(initial_actions),
  ncol=length(actions)
)

for(i in seq_len(learning_starts)){
  replay_add(
    training_buffer_test,
    obs=dummy_obs,
    action_onehot=dummy_action,
    rewards=rep(0,length(initial_actions)),
    next_obs=dummy_obs,
    masks=dummy_masks,
    next_masks=dummy_masks,
    done=0
  )
}

training_update_test<-maddpg_train_step(
  training_system_test,
  training_buffer_test
)

stopifnot(nrow(training_update_test$log)==length(initial_actions))
stopifnot(all(is.finite(training_update_test$log$actor_loss)))
stopifnot(all(is.finite(training_update_test$log$critic_loss)))
```

# . Convert combined herd pressure to an ecological footprint

If several agents occupy the same cell, their herd sizes are added before the ecological transition is sampled to account for the total grazing pressure in each cell. 

```{r combined-pressure}
herd_size_to_action<-function(total_herd_size) {
  case_when(
    total_herd_size <= 0 ~ "none",
    total_herd_size < 200 ~ "<200",
    total_herd_size < 400 ~ "200-400",
    total_herd_size < 600 ~ "400-600",
    TRUE ~ ">600"
  )
}
```

```{r test-combined-pressure, include=FALSE}
stopifnot(
  herd_size_to_action(0) == "none",
  herd_size_to_action(100) == "<200",
  herd_size_to_action(300) == "200-400",
  herd_size_to_action(500) == "400-600",
  herd_size_to_action(700) == ">600",
  herd_size_to_action(1400) == ">600"
)
```

# . Update the landscape once per step

Each of the 25 cells is updated once per iteration and keeps its own persistent ecological EMA. For grazed cells, the stochastic ecological transition uses the **combined** herd pressure of all agents present. For ungrazed cells, ecological footprint is `none`: grazing pressure is zero, but weather still affects the stochastic transition outcome. A previously grazed emergency cell therefore retains its ecological history after the herder leaves. Every update is written to the cell-level ledger.

```{r landscape-update}
update_landscape<-function(land_grid,action_log,transition_weather,step){
  pressure_by_cell<-action_log%>%
    group_by(cell_id)%>%
    summarise(total_herd_size=sum(herd_size),n_agents=n(),.groups="drop")%>%
    mutate(ecological_footprint=map_chr(total_herd_size,herd_size_to_action))
  new_grid<-land_grid
  ledger_list<-vector("list",nrow(land_grid))
  for(j in seq_len(nrow(land_grid))){
    cell<-land_grid$cell_id[j]
    current_condition<-land_grid$rangeland_condition[j]
    old_ema<-land_grid$condition_ema_score[j]
    pressure<-pressure_by_cell%>%filter(.data$cell_id==cell)
    if(nrow(pressure)==0){
      ecological_footprint<-"none"
      total_herd_size<-0
      n_agents<-0
    }else{
      ecological_footprint<-pressure$ecological_footprint[[1]]
      total_herd_size<-pressure$total_herd_size[[1]]
      n_agents<-pressure$n_agents[[1]]
    }
    transition_outcome<-sample_next_condition(
      current_condition=current_condition,
      action=ecological_footprint,
      weather=transition_weather
    )
    memory_update<-update_cell_memory(old_ema=old_ema,transition_outcome=transition_outcome)
    new_grid$condition_ema_score[j]<-memory_update$ema_score
    new_grid$rangeland_condition[j]<-memory_update$condition
    ledger_list[[j]]<-tibble(
      step=step,
      from_time=step,
      to_time=step+1L,
      cell_id=cell,
      transition_weather=transition_weather,
      condition_before=current_condition,
      ecological_footprint=ecological_footprint,
      transition_outcome=transition_outcome,
      ema_before=old_ema,
      ema_after=memory_update$ema_score,
      condition_after=memory_update$condition,
      total_herd_size=total_herd_size,
      n_agents=n_agents
    )
  }
  list(land_grid=new_grid,ledger=bind_rows(ledger_list))
}
```

```{r test-landscape-update, include=FALSE}
test_grid<-land_grid_initial
test_grid$condition_ema_score[test_grid$cell_id==2]<-1.8
test_grid$rangeland_condition[test_grid$cell_id==2]<-"poor"
test_actions<-tibble(
  agent_id=c("A1","A2"),
  cell_id=c(1L,1L),
  action=c(">600",">600"),
  herd_size=c(700,700)
)
set.seed(101)
land_update_test<-update_landscape(
  land_grid=test_grid,
  action_log=test_actions,
  transition_weather="drought",
  step=1L
)
stopifnot(nrow(land_update_test$ledger)==25)
stopifnot(land_update_test$ledger$ecological_footprint[land_update_test$ledger$cell_id==1]==">600")
stopifnot(land_update_test$ledger$total_herd_size[land_update_test$ledger$cell_id==1]==1400)
stopifnot(land_update_test$ledger$ecological_footprint[land_update_test$ledger$cell_id==2]=="none")
stopifnot(is.finite(land_update_test$land_grid$condition_ema_score[land_update_test$land_grid$cell_id==2]))
stopifnot(land_update_test$ledger$ema_before[land_update_test$ledger$cell_id==2]==1.8)
stopifnot(
  land_update_test$land_grid$condition_ema_score[land_update_test$land_grid$cell_id==2]==
  land_update_test$ledger$ema_after[land_update_test$ledger$cell_id==2]
)
```

# . One simultaneous MARL step

This is the core multi-agent step.

Order of operations:

1. Resolve each agent's current location using the current EMA-defined landscape and current weather.
2. Share current locations and previous stocking-category herd sizes.
3. Construct each agent's local numeric observation.
4. Each decentralised actor selects a feasible discrete stocking action using masked straight-through Gumbel-Softmax.
5. Calculate shared-cell interaction counts and immediate Action-PES rewards.
6. Update every cell from \(t\) to \(t+1\) using the unchanged ecological transition model.
7. Resolve next locations and construct next observations.
8. Store the joint transition \((O_t,A_t,R_t,O_{t+1})\) in replay memory.
9. After the replay warm-up, update all centralised critics and decentralised actors from a random mini-batch.
10. Polyak-update target actor and critic networks.
11. Carry each realised discrete stocking action forward as the next previous-action anchor.
12. Write agent, reward, ecological and learning ledgers.

```{r marl-step}
marl_step<-function(
    agents,
    land_grid,
    weather,
    next_weather,
    step,
    maddpg,
    replay_buffer,
    train=TRUE,
    explore=TRUE
){
  # ------------------------------------------------------------
  # A. Resolve locations in current state
  # ------------------------------------------------------------
  location_results<-map(
    names(agents),
    ~resolve_location(
      agent_id=.x,
      step=step,
      weather=weather,
      land_grid=land_grid
    )
  )
  names(location_results)<-names(agents)

  for(id in names(agents)){
    agents[[id]]$current_cell<-location_results[[id]]$cell_id
  }

  # ------------------------------------------------------------
  # B. Shared information before action selection
  # ------------------------------------------------------------
  shared_info<-share_agent_information(
    agents,
    land_grid
  )

  # ------------------------------------------------------------
  # C. Local observations o_i,t
  # ------------------------------------------------------------
  state_keys<-setNames(
    map_chr(
      agents,
      observe_agent_state,
      land_grid=land_grid,
      weather=weather,
      shared_info=shared_info
    ),
    names(agents)
  )

  observation_list<-map(
    agents,
    observe_agent_vector,
    land_grid=land_grid,
    weather=weather,
    shared_info=shared_info
  )

  observations<-do.call(
    rbind,
    observation_list
  )
  rownames(observations)<-names(agents)

  current_masks<-do.call(
    rbind,
    lapply(
      agents,
      function(agent)action_mask(agent$previous_action)
    )
  )
  rownames(current_masks)<-names(agents)

  # ------------------------------------------------------------
  # D. Decentralised actor choice with Gumbel-Softmax
  # ------------------------------------------------------------
  action_results<-setNames(
    lapply(
      seq_along(agents),
      function(i){
        id<-names(agents)[[i]]
        select_action_actor(
          actor=maddpg$actors[[id]],
          observation=observations[i,],
          previous_action=agents[[id]]$previous_action,
          explore=explore
        )
      }
    ),
    names(agents)
  )

  chosen_actions<-setNames(
    map_chr(action_results,"action"),
    names(agents)
  )

  action_onehot<-do.call(
    rbind,
    lapply(action_results,`[[`,"one_hot")
  )
  rownames(action_onehot)<-names(agents)

  herd_sizes<-action_herd_size[chosen_actions]

  # ------------------------------------------------------------
  # E. Simultaneous action log
  # ------------------------------------------------------------
  action_log<-tibble(
    agent_id=names(agents),
    cell_id=map_int(agents,"current_cell"),
    previous_action=map_chr(agents,"previous_action"),
    action=unname(chosen_actions),
    herd_size=as.numeric(herd_sizes),
    state_key=unname(state_keys),
    feasible_actions=map_chr(
      action_results,
      ~paste(.x$feasible_actions,collapse=" | ")
    ),
    chosen_probability=map2_dbl(
      action_results,
      chosen_actions,
      ~unname(.x$probabilities[[.y]])
    ),
    movement_mode=map_chr(location_results,"movement_mode"),
    emergency_cost=map_dbl(location_results,"emergency_cost"),
    travel_cost=map_dbl(location_results,"travel_cost"),
    social_cost=map_dbl(location_results,"social_cost")
  )%>%
    group_by(cell_id)%>%
    mutate(n_agents_same_cell=n())%>%
    ungroup()

  # ------------------------------------------------------------
  # F. Immediate Action-PES reward
  # ------------------------------------------------------------
  action_log<-action_log%>%
    left_join(
      land_grid%>%
        select(cell_id,rangeland_condition),
      by="cell_id"
    )

  reward_components<-pmap(
    list(
      action_log$action,
      action_log$rangeland_condition,
      action_log$n_agents_same_cell,
      action_log$emergency_cost
    ),
    function(
        action,
        rangeland_condition,
        n_agents_same_cell,
        emergency_cost
    ){
      compute_reward(
        action=action,
        rangeland_condition=rangeland_condition,
        n_agents_same_cell=n_agents_same_cell,
        emergency_cost=emergency_cost
      )
    }
  )

  action_log$reward<-map_dbl(reward_components,"net_reward")
  action_log$livestock_income<-map_dbl(reward_components,"livestock_income")
  action_log$pes_reward<-map_dbl(reward_components,"pes_reward")
  action_log$gross_reward<-map_dbl(reward_components,"gross_reward")
  action_log$stock_survival<-map_dbl(reward_components,"stock_survival")
  action_log$expected_surviving_herd<-map_dbl(reward_components,"expected_surviving_herd")
  action_log$expected_stock_loss<-map_dbl(reward_components,"expected_stock_loss")
  action_log$interaction_multiplier<-map_dbl(reward_components,"interaction_multiplier")

  # ------------------------------------------------------------
  # G. Ecology evolves t -> t+1
  # ------------------------------------------------------------
  landscape_update<-update_landscape(
    land_grid=land_grid,
    action_log=action_log,
    transition_weather=next_weather,
    step=step
  )

  next_land_grid<-landscape_update$land_grid

  # ------------------------------------------------------------
  # H. Construct next agent locations and observations
  # ------------------------------------------------------------
  next_agents<-agents

  for(id in names(next_agents)){
    next_agents[[id]]$previous_action<-chosen_actions[[id]]
    next_agents[[id]]$herd_size<-unname(
      action_herd_size[[chosen_actions[[id]]]]
    )

    next_location<-resolve_location(
      agent_id=id,
      step=step+1L,
      weather=next_weather,
      land_grid=next_land_grid
    )

    next_agents[[id]]$current_cell<-next_location$cell_id
  }

  next_shared_info<-share_agent_information(
    next_agents,
    next_land_grid
  )

  next_state_keys<-setNames(
    map_chr(
      next_agents,
      observe_agent_state,
      land_grid=next_land_grid,
      weather=next_weather,
      shared_info=next_shared_info
    ),
    names(next_agents)
  )

  next_observation_list<-map(
    next_agents,
    observe_agent_vector,
    land_grid=next_land_grid,
    weather=next_weather,
    shared_info=next_shared_info
  )

  next_observations<-do.call(
    rbind,
    next_observation_list
  )
  rownames(next_observations)<-names(next_agents)

  next_masks<-do.call(
    rbind,
    lapply(
      next_agents,
      function(agent)action_mask(agent$previous_action)
    )
  )
  rownames(next_masks)<-names(next_agents)

  # ------------------------------------------------------------
  # I. Store one joint transition in replay memory
  # ------------------------------------------------------------
  reward_vector<-setNames(
    action_log$reward,
    action_log$agent_id
  )[names(agents)]

  replay_add(
    replay_buffer,
    obs=observations,
    action_onehot=action_onehot,
    rewards=reward_vector,
    next_obs=next_observations,
    masks=current_masks,
    next_masks=next_masks,
    done=0
  )

  # ------------------------------------------------------------
  # J. MADDPG gradient updates
  # ------------------------------------------------------------
  training_rows<-list()

  if(train&&replay_buffer$size>=learning_starts){
    for(u in seq_len(updates_per_step)){
      update<-maddpg_train_step(
        maddpg,
        replay_buffer
      )
      maddpg<-update$system
      training_rows[[u]]<-update$log
    }
  }

  training_log<-bind_rows(training_rows)

  if(nrow(training_log)>0){
    training_log<-training_log%>%
      mutate(step=step)
  }

  action_log<-action_log%>%
    mutate(
      step=step,
      weather=weather,
      next_weather=next_weather
    )

  list(
    agents=next_agents,
    maddpg=maddpg,
    replay_buffer=replay_buffer,
    land_grid=next_land_grid,
    action_log=action_log,
    cell_ledger=landscape_update$ledger,
    training_log=training_log,
    state_keys=state_keys,
    next_state_keys=next_state_keys,
    shared_info=shared_info,
    next_shared_info=next_shared_info
  )
}
```

```{r test-marl-step, include=FALSE}
set.seed(101)
torch_manual_seed(101)

one_step_system<-create_maddpg_system()
one_step_buffer<-new_replay_buffer()

one_step_test<-marl_step(
  agents=agents_initial,
  land_grid=land_grid_initial,
  weather="normal",
  next_weather="drought",
  step=1,
  maddpg=one_step_system,
  replay_buffer=one_step_buffer,
  train=TRUE,
  explore=TRUE
)

stopifnot(nrow(one_step_test$action_log)==6)
stopifnot(nrow(one_step_test$cell_ledger)==25)
stopifnot(one_step_test$replay_buffer$size==1L)

# Every realised action respects the adjacent-category constraint.
stopifnot(all(
  map2_lgl(
    one_step_test$action_log$previous_action,
    one_step_test$action_log$action,
    ~.y%in%allowed_actions(.x)
  )
))

# Ecology uses next weather to generate the next condition.
stopifnot(all(
  one_step_test$cell_ledger$transition_weather=="drought"
))
```

# . Full MADDPG simulation

The weather sequence has length `steps + 1` so that \(s'\) includes the next observed weather condition when the replay transition is stored and MADDPG is updated.

```{r simulate-marl}
simulate_marl<-function(
    steps=300L,
    seed=NULL,
    initial_agents=NULL,
    initial_land_grid=NULL,
    initial_maddpg=NULL,
    train=TRUE,
    explore=TRUE,
    return_maddpg=FALSE
){
  if(!is.null(seed)){
    set.seed(seed)
    torch_manual_seed(as.integer(seed))
  }

  agents<-if(is.null(initial_agents)){
    setNames(
      lapply(
        names(initial_actions),
        function(id){
          create_agent(
            agent_id=id,
            initial_action=initial_actions[[id]]
          )
        }
      ),
      names(initial_actions)
    )
  }else{
    initial_agents
  }

  land_grid<-if(is.null(initial_land_grid)){
    make_land_grid()
  }else{
    initial_land_grid
  }

  maddpg<-if(is.null(initial_maddpg)){
    create_maddpg_system(
      agent_ids=names(agents)
    )
  }else{
    initial_maddpg
  }

  replay_buffer<-new_replay_buffer()

  # W_t is observed in o_t. W_(t+1) helps generate C_(t+1).
  weather_sequence<-draw_weather(steps+1L)

  action_logs<-vector("list",steps)
  cell_ledgers<-vector("list",steps)
  training_logs<-vector("list",steps)
  shared_ledgers<-vector("list",steps)
  landscape_counts<-vector("list",steps)

  for(t in seq_len(steps)){
    result<-marl_step(
      agents=agents,
      land_grid=land_grid,
      weather=weather_sequence[t],
      next_weather=weather_sequence[t+1L],
      step=t,
      maddpg=maddpg,
      replay_buffer=replay_buffer,
      train=train,
      explore=explore
    )

    agents<-result$agents
    maddpg<-result$maddpg
    replay_buffer<-result$replay_buffer
    land_grid<-result$land_grid

    action_logs[[t]]<-result$action_log
    cell_ledgers[[t]]<-result$cell_ledger
    training_logs[[t]]<-result$training_log
    shared_ledgers[[t]]<-result$shared_info%>%
      mutate(step=t)

    condition_count<-table(
      factor(
        land_grid$rangeland_condition,
        levels=r_states
      )
    )

    landscape_counts[[t]]<-tibble(
      step=t,
      intact=unname(condition_count["intact"]),
      good=unname(condition_count["good"]),
      poor=unname(condition_count["poor"]),
      degraded=unname(condition_count["degraded"]),
      pct_degraded=
        100*unname(condition_count["degraded"])/nrow(land_grid),
      mean_ema=mean(land_grid$condition_ema_score),
      max_ema=max(land_grid$condition_ema_score)
    )
  }

  list(
    agents=agents,
    maddpg=if(return_maddpg)maddpg else NULL,
    replay_size=replay_buffer$size,
    land_grid=land_grid,
    weather=weather_sequence,
    action_log=bind_rows(action_logs),
    cell_ledger=bind_rows(cell_ledgers),
    training_log=bind_rows(training_logs),
    shared_ledger=bind_rows(shared_ledgers),
    landscape_counts=bind_rows(landscape_counts)
  )
}
```

```{r test-simulation, include=FALSE}
simulation_test<-simulate_marl(
  steps=100L,
  seed=321L,
  train=TRUE,
  explore=TRUE
)

stopifnot(nrow(simulation_test$action_log)==100*6)
stopifnot(nrow(simulation_test$cell_ledger)==100*25)
stopifnot(simulation_test$replay_size==100L)
stopifnot(nrow(simulation_test$training_log)>0)

stopifnot(all(
  simulation_test$landscape_counts$degraded>=0&
    simulation_test$landscape_counts$degraded<=25
))

# The environment must generate stochastic transition outcomes.
stopifnot(any(
  simulation_test$cell_ledger$transition_outcome!=
    simulation_test$cell_ledger$condition_before
))

# Ungrazed cells retain ecological-memory updates.
ungrazed_rows<-simulation_test$cell_ledger%>%
  filter(ecological_footprint=="none")

stopifnot(nrow(ungrazed_rows)>0)
stopifnot(all(is.finite(ungrazed_rows$ema_after)))

# Every realised herder action obeys the adjacent-category constraint.
stopifnot(all(
  map2_lgl(
    simulation_test$action_log$previous_action,
    simulation_test$action_log$action,
    ~.y%in%allowed_actions(.x)
  )
))

simulation_test$landscape_counts%>%
  select(step,intact,good,poor,degraded,mean_ema,max_ema)%>%
  tail()
```

# . Train one MADDPG system

This is one continuous learning trajectory. Actor and critic parameters, target networks and replay memory carry forward throughout the run.

```{r train-model}
results_train<-simulate_marl(
  steps=500L,
  seed=128L,
  train=TRUE,
  explore=TRUE,
  return_maddpg=TRUE
)
```

# . Core diagnostics

## .1 Landscape degradation

```{r plot-degradation}
p_degraded_pct<-results_train$landscape_counts %>%
  ggplot(aes(step, pct_degraded)) +
  geom_line() +
  labs(
    title = "Percentage of degraded cells",
    x = "Step",
    y = "Degraded cells (%)"
  ) +
  theme_minimal()
p_degraded_pct
```

## .2 Action mix through time

```{r plot-action-mix}
results_train$action_log %>%
  mutate(
    period = cut(
      step,
      breaks = seq(0,500, by = 50),
      include.lowest = TRUE
    )
  ) %>%
  count(period, action) %>%
  group_by(period) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(period, prop, fill = action)) +
  geom_col() +
  labs(
    title = "Learned stocking-action mix through time",
    x = "Training period",
    y = "Proportion of choices",
    fill = "Action"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## .3 Emergency movements

```{r plot-emergency}
results_train$action_log %>%
  count(step, movement_mode) %>%
  ggplot(aes(step, n, fill = movement_mode)) +
  geom_col() +
  labs(
    title = "Emergency and designated-pasture movement",
    x = "Step",
    y = "Number of agents",
    fill = "Movement mode"
  ) +
  theme_minimal()
```

## .4 MADDPG learning losses

```{r plot-maddpg-losses}
results_train$training_log%>%
  pivot_longer(
    cols=c(actor_loss,critic_loss),
    names_to="loss_type",
    values_to="loss"
  )%>%
  mutate(
    period=ceiling(step/25)
  )%>%
  group_by(
    period,
    agent_id,
    loss_type
  )%>%
  summarise(
    mean_loss=mean(loss),
    .groups="drop"
  )%>%
  ggplot(
    aes(
      period,
      mean_loss,
      colour=loss_type
    )
  )+
  geom_line()+
  facet_wrap(~agent_id,scales="free_y")+
  labs(
    title="MADDPG actor and critic losses through learning",
    x="25-step period",
    y="Mean loss",
    colour="Loss"
  )+
  theme_minimal()
```

## .5 Cell-level ecological ledger

This shows more details of the ecological dynamics, an insight that illustrates what is going on inside the aggregated degradation percentage.

```{r plot-cell-ledger}
condition_numeric<-c(
  intact = 0,
  good = 1,
  poor = 2,
  degraded = 3
)
results_train$cell_ledger %>%
  mutate(
    condition_value =
      condition_numeric[condition_after]
  ) %>%
  ggplot(
    aes(
      step,
      factor(cell_id),
      fill = condition_value
    )
  ) +
  geom_tile() +
  scale_fill_viridis_c(
    breaks = 0:3,
    labels = r_states
  ) +
  labs(
    title = "Cell-level ecological ledger",
    x = "Step",
    y = "Cell ID",
    fill = "Condition"
  ) +
  theme_minimal()
```

# . Independent replicate workflow

We need a continuous training trajectory with independent simulation replicates.This function creates independent runs by resetting both:

- the ecological landscape;
- all actor and critic networks, target networks and replay memory.

```{r independent-replicates}
get_parallel_workers<-function(max_workers=6L,reserve=2L){
  available<-as.integer(future::availableCores())
  as.integer(
    max(
      1L,
      min(max_workers,max(1L,available-reserve))
    )
  )
}

run_independent_replicates<-function(
    n_runs=20L,
    steps=250L,
    seed_start=1000L,
    workers=NULL
){
  if(!requireNamespace("future",quietly=TRUE)||
     !requireNamespace("furrr",quietly=TRUE)){
    stop(
      "Parallel runs require packages 'future' and 'furrr'. ",
      "Install once with install.packages(c('future','furrr'))."
    )
  }

  if(is.null(workers)){
    workers<-get_parallel_workers()
  }

  workers<-as.integer(max(1L,workers))

  message(
    "Running ",
    n_runs,
    " independent MADDPG simulations on ",
    workers,
    " workers."
  )

  future::plan(
    future::multisession,
    workers=workers
  )

  on.exit(
    future::plan(future::sequential),
    add=TRUE
  )

  furrr::future_map_dfr(
    seq_len(n_runs),
    function(run_id){
      # Each replicate constructs new actors, critics, target networks
      # and a new replay buffer inside simulate_marl().
      res<-simulate_marl(
        steps=steps,
        seed=seed_start+run_id,
        train=TRUE,
        explore=TRUE,
        return_maddpg=FALSE
      )

      res$landscape_counts%>%
        mutate(run=run_id)
    },
    .options=furrr::furrr_options(
      seed=TRUE,
      scheduling=1,
      packages=c("torch","tidyverse","purrr")
    )
  )
}

replicate_degradation<-run_independent_replicates(
  n_runs=20L,
  steps=500L,
  seed_start=2000L
)
```

```{r test-independent-replicates, include=FALSE}
stopifnot(
  n_distinct(
    replicate_degradation$run
  ) == 20
)
replicate_degradation %>%
  ggplot(
    aes(
      step,
      degraded,
      group = run
    )
  ) +
  geom_line(alpha = 0.2) +
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  labs(
    title = "TEST: independent runs are done correctly",
    y = "Number of degraded cells"
  ) +
  theme_minimal()
```
```{r action mix visualisation, include=FALSE}
# ============================================================
# ACTION MIX UNDER DIFFERENT PES LEVELS
# ============================================================

action_order<-c("<200","200-400","400-600",">600")

action_cols<-c(
  "<200"="#F3B6C6",
  "200-400"="#D77FB3",
  "400-600"="#A64F9E",
  ">600"="#5B2A86"
)

base_pes_payment<-pes_payment

run_pes_level<-function(multiplier,steps=500L,seed=128L){
  pes_payment<<-base_pes_payment*multiplier

  res<-simulate_marl(
    steps=steps,
    seed=seed,
    train=TRUE,
    explore=TRUE
  )

  res$action_log%>%
    mutate(pes_multiplier=multiplier)
}
#A small PES grid
pes_action_runs<-map_dfr(
  c(0.5,1,1.5),
  run_pes_level
)

pes_payment<-base_pes_payment

action_mix_pes<-pes_action_runs%>%
  mutate(
    setting=case_when(
      pes_multiplier==0.5~"Low PES",
      pes_multiplier==1~"Baseline",
      pes_multiplier==1.5~"High PES"
    ),
    setting=factor(setting,levels=c("Low PES","Baseline","High PES")),
    action=factor(action,levels=action_order)
  )%>%
  count(setting,step,action,name="n")%>%
  group_by(setting,step)%>%
  mutate(prop=n/sum(n))%>%
  ungroup()

p_action_pes<-ggplot(action_mix_pes,aes(step,prop,fill=action))+
  geom_area()+
  facet_grid(setting~.)+
  scale_fill_manual(values=action_cols,drop=FALSE)+
  scale_y_continuous(
    limits=c(0,1),
    labels=scales::percent_format(accuracy=1)
  )+
  labs(
    x="Iteration",
    y="Proportion of actions",
    fill="Stocking action",
    title="MADDPG action mix under different PES levels"
  )+
  theme_minimal()

p_action_pes
```

